A mixture of hidden Markov models to predict the lymphatic spread in head and neck cancer

### 2024-01-17 content/manuscript.qmd
add more co-authors

---
title: A mixture of hidden Markov models to predict the lymphatic spread in head and neck cancer
author:
  - name: Roman Ludwig
    orcid: 0000-0001-9434-328X
    email: roman.ludwig@usz.ch
    affiliations:
      - ref: 1
      - ref: 2
  - name: Julian Brönnimann
    affiliations:
      - ref: 1
      - ref: 2
  - name: Jan Unkelbach
    orcid: 0000-0002-4275-990X
    affiliations:
      - ref: 1
      - ref: 2
affiliations:
  - id: 1
    name: University of Zurich
    departement: Departement of Physics
    city: Zurich
    country: Switzerland
  - id: 2
    name: University Hospital Zurich
    departement: Departement of Radiation Oncology
    city: Zurich
    country: Switzerland
bibliography: ../references.bib
abstract: |
  We previously developed a mechanistic hidden Markov model (HMM) to predict the
  lymphatic tumor progression in oropharyngeal squamous cell carcinomas (OPSCCs). To
  extend the model to other tumor subsites, defined by ICD-10 codes, in the head and
  neck, we employ a mixture of these HMMs and learn the cluster assignments and model
  parameters in an iterative, EM-like algorithm from multi-centric data. The mixture
  model manages to cluster anatomically close subsites and correctly infers the
  clusters' parameters. Using this mixture model allows the prediction of individual
  risks of occult nodal disease, given a diagnosis that includes tumor subsite.
format:
  iccr-pdf:
    keep-tex: true
    include-in-header:
      - text: "\\usepackage{newtxtext,newtxmath}"
  html:
    code-line-numbers: true
    code-tools: true
crossref:
  fig-prefix: "figure"
  tbl-prefix: "table"
  eq-prefix: "equation"
  sec-prefix: "section"
---

@@

![Parametrized graph representation of the lymphatic network considering four LNLs. Blue, round nodes represent the hidden RVs, while orange square nodes show the observed RVs. Arcs represent a conditional probability parametrized with the quantity noted next to it.](../static/graph_with_obs.svg){#fig-graph-with-obs width="22%"}

@@

When a patient gets diagnosed, their LNLs' involvement status is assessed using a clinical modality $\mathcal{O}$ with its repsective sensitivity $s_N^\mathcal{O}$ and specificity $s_P^\mathcal{O}$. This involvement status, too, is modelled as an RV $Y_v^\mathcal{O}$. Hence, a diagnosis obtained with diagnostic modality $\mathcal{O}$ can be collected in the random vector $\mathbf{Y}^\mathcal{O}$.

We model the process of tumor progression via lymphatic drainage by connecting the hidden RVs in a graph, as shown in @fig-graph-with-obs. The arcs in [this graph](#fig-graph-with-obs) represent conditional probabilities. The orange arcs, as mentioned, correspond to observing a diagnosis $Z_v^\mathcal{O}$, given a hidden state $X_v$. The values this conditional probability $P(Y_v \mid X_v)$ can take on are tabulated in @tbl-sens-spec.

>>>>

---
title: A mixture of hidden Markov models to predict the lymphatic spread in head and neck cancer
author:
  - name: Roman Ludwig
    orcid: 0000-0001-9434-328X
    email: roman.ludwig@usz.ch
    affiliations:
      - ref: 1
      - ref: 2
  - name: Julian Brönnimann
    affiliations:
      - ref: 1
      - ref: 2
  - name: Yoel Perez Haas
    affiliations:
      - ref: 1
      - ref: 2
  - name: Esmée Lauren Looman
    affiliations:
      - ref: 1
      - ref: 2
  - name: Adrian Schubert
    affiliations:
      - ref: 3
      - ref: 4
      - ref: 7
  - name: Dorothea Barbatei
    affiliations:
      - ref: 8
  - name: Laurence Bauwens
    affiliations:
      - ref: 8
  - name: Jean-Marc Hoffmann
    orcid: 0000-0003-1193-0863
    affiliations:
      - ref: 2
  - name: Sandrine Werlen
    affiliations:
      - ref: 4
      - ref: 5
  - name: Olgun Elicin
    orcid: 0000-0002-6996-0646
    affiliations:
      - ref: 3
  - name: Matthias Dettmer
    orcid: 0000-0003-0948-1392
    affiliations:
      - ref: 6
      - ref: 10
  - name: Philippe Zrounba
    orcid: 0000-0001-6223-6113
    affiliations:
      - ref: 9
  - name: Bertrand Poumayou
    orcid: 0000-0002-8060-7827
    affiliations:
      - ref: 2
  - name: Panagiotis Balermpas
    orcid: 0000-0001-5261-6446
    affiliations:
      - ref: 2
  - name: Vincent Grégoire
    affiliations:
      - ref: 8
  - name: Roland Giger
    orcid: 0000-0002-5574-3210
    affiliations:
      - ref: 4
      - ref: 5
  - name: Jan Unkelbach
    orcid: 0000-0002-4275-990X
    affiliations:
      - ref: 1
      - ref: 2
affiliations:
  - id: 1
    name: University of Zurich
    department: Department of Physics
    city: Zurich
    country: Switzerland
  - id: 2
    name: University Hospital Zurich
    department: Department of Radiation Oncology
    city: Zurich
    country: Switzerland
  - id: 3
    name: Bern University Hospital
    department: Department of Radiation Oncology
    city: Bern
    country: Switzerland
  - id: 4
    name: Bern University Hospital
    department: Department of ENT, Head & Neck Surgery
    city: Bern
    country: Switzerland
  - id: 5
    name: Bern University Hospital
    department: Head and Neck Anticancer Center
    city: Bern
    country: Switzerland
  - id: 6
    name: Bern University Hospital
    department: Institute of Tissue Medicine and Pathology
    city: Bern
    country: Switzerland
  - id: 7
    name: Réseau Hospitalier Neuchâtelois
    department: Department of ENT, Head & Neck Surgery
    city: Neuchâtelois
    country: Switzerland
  - id: 8
    name: Centre Léon Bérard
    department: Department of Radiation Oncology
    city: Lyon
    country: France
  - id: 9
    name: Centre Léon Bérard
    department: Department of Head and Neck Surgery
    city: Lyon
    country: France
  - id: 10
    name: Klinikum Stuttgart
    department: Institute of Pathology
    city: Stuttgart
    country: Germany
bibliography: ../references.bib
abstract: |
  We previously developed a mechanistic hidden Markov model (HMM) to predict the
  lymphatic tumor progression in oropharyngeal squamous cell carcinomas (OPSCCs). To
  extend the model to other tumor subsites, defined by ICD-10 codes, in the head and
  neck, we employ a mixture of these HMMs and learn the cluster assignments and model
  parameters in an iterative, EM-like algorithm from multi-centric data. The mixture
  model manages to cluster anatomically close subsites and correctly infers the
  clusters' parameters. Using this mixture model allows the prediction of individual
  risks of occult nodal disease, given a diagnosis that includes tumor subsite.
format:
  iccr-pdf:
    keep-tex: true
    include-in-header:
      - text: "\\usepackage{newtxtext,newtxmath}"
  html:
    code-line-numbers: true
    code-tools: true
crossref:
  fig-prefix: "figure"
  tbl-prefix: "table"
  eq-prefix: "equation"
  sec-prefix: "section"
---

@@

![Parametrized graph representation of the lymphatic network considering four LNLs. Blue, round nodes represent the hidden RVs, while orange square nodes show the observed RVs. Arcs represent a conditional probability parametrized with the quantity noted next to it.](../static/graph_with_obs.svg){#fig-graph-with-obs width="20%"}

@@

When a patient gets diagnosed, their LNLs' involvement status is assessed using a clinical modality (e.g. computed tomography (CT) or magnetic resonance imaging (MRI)) with its repsective sensitivity $s_N$ and specificity $s_P$. This involvement status, too, is modelled as an RV $Y_v$. Hence, a diagnosis obtained with this diagnostic modality can be collected in the random vector $\mathbf{Y}$.

We model the process of tumor progression via lymphatic drainage by connecting the hidden RVs in a graph, as shown in @fig-graph-with-obs. The arcs in [this graph](#fig-graph-with-obs) represent conditional probabilities. The orange arcs, as mentioned, correspond to observing a diagnosis $Y_v$, given a hidden state $X_v$. The values this conditional probability $P(Y_v \mid X_v)$ can take on are tabulated in @tbl-sens-spec.
